\section{Conclusions and Future Research}

We have seen ideas from sorting can be modified to fit Quickhull as well.
We obtain a $1.5$~\textendash~$13\times$ sequentially and a 
$1.3$~\textendash~$8.6\times$ speedup in parallel. Our implementation performs
best on an instruction set supporting SIMD compression instructions.

We obtain roughly $80\%$ of our architecture's bandwidth, so to obtain 
significant improvements future work should focus on finding a convex hull
algorithm with lower bandwidth requirements. A step in this direction could
be to efficiently vectorize and parallelise the Akl-Toussaint heuristic as
described in Section~\ref{sec:evaluation}.
As Quickhull resembles Quicksort, this heuristic resembles Samplesort.
It may be possible to adept the branchless implementation of ips4o \cite{ips4o}.

We have verified our output with the output of competing implementations, but
discussion about the accuracy of Quickhull is missing and should be addressed
by future work.

\tkcomment{Move to related work?}
As a baseline we have considered Qhull \cite{}, the algorithms of CGAL \cite{},
and the implementation of PBBS. Some preliminary testing shows that the
implementation of PBBS is the fastest by a large margin. In
Table~\ref{table:reference} are the runtimes for a disk of $10^7$ points.

\begin{table}[ht]
    \caption{Runtime in seconds for a disk of $10^7$ points}
    \label{table:reference}
    \begin{tabular}{c | c }
     Implementation & Runtime \\ 
     \hline \\
     PBBS & 0.31 \\  
     CGAL Quickhull & 0.61 \\
     CGAL Akl-Toussaint & 0.60 \\
     CGAL Bykat & 0.73 \\
     CGAL Eddy & 0.98 \\
     CGAL Graham & 1.3 \\
     CGAL Jarvis & 208 \\
     Qhull & 1.1 \\
    \end{tabular}
\end{table}


